Combinatorial interpretations of holonomic recurrences

### 2021-08-29 article/main.tex
Make minor fixes

Every recursive structure can be described using a generating function, which can be then transformed into holonomic recurrence. It allows us to quickly count objects of given structure with of given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations, like adding some vertices to the smaller trees. Although, there is no known method of interpreting it in the language of the combinatorial structures.

@@

In the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an approach that helped us to find an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation.

@@

Section\ \ref{sec:brutal_iterative_approach} presents our final approach to the original problem, ie.\ computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is able to provide us with the full solution, but we were not able to achieve that due to the limited time that was available to us.

@@

In order to make reading this article easier, let's fix some notation and make some assumptions. Any deviation from this will be clearly stated. If you have trouble guessing what some letter or word means, this should be your primary place to look for the answer.

@@

\(\no\) is implicitly defined as an identifier of a non-existing node (eg.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

Dashed edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__dashed_edge}
\end{minipage}
depict edges that may or may not exist (eg.\ when we describe transformations and we don't care if the node is root or if it has children).

@@

Virtual node is an artificially defined node that does not affect the structure definition (eg.\ we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

@@

eg.\ stands for Latin exempli gratia, which means \textit{for example}.

ie.\ stands for Latin id est, which means \textit{that is}.

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define a full binary tree recursively as a node having no children (ie.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

In the article we use ordinary generating functions to describe and count structures.

@@

The highlighted terms are describing following tree:

@@

\begin{definition}
    \label{def:holo_1}
    \cite{holotoolkit}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) exists holonomic recurrence for it, ie.:

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, ie. none of them satisfies a linear recurrence equation with polynomial coefficients\cite{nonholo}.
\end{example}

\begin{definition}
    \label{def:holo_2}
    \cite{complexity}
    Equivalently to definition\ \ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) there exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identical to zero.
\end{definition}

@@

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, ie.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, eg.\ we may want to find the number of trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent.

@@

In the article we try to gather methods and results that may be helpful for anyone trying to find such interpretation, as we believe that such interpretation may give better insight into the nature of the structure and make it easier to study its behaviour.

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as:
    \[\enc{T} = \enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\]
    ie.\ catenation of the identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

@@

In the article we define some operations using its notation. As pre-order notation is bijective, you can obtain the structure the notation describes by reversing the process of encoding.

@@

We define the size of a tree as a number of internal nodes (ie.\ nodes that are not leaves). Our generating function is therefore:

@@

It is easy to see that a tree of size \(n\) (ie. with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

Let's define two local operations: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of tree with predefined size (eg.\ node and its 2 children, but can not traverse the whole tree).

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), ie. trees of size \(n\) with a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\)

@@

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary trees. We don't change the structure, just the notation.

@@

The bijection between unary-binary and left-leaning binary trees is very simple. We just represent unary nodes as nodes with just the left child of the left-leaning binary tree. The other cases translate immediately to corresponding ones. Please note that this transformation keeps the size of the tree (ie.~number of edges).

@@

All we need to show are the transformations from the smaller cases, ie. \(U(n-1)\) and \(U(n-2)\).

@@

This definition is quite precise, but is difficult to implement as a structural recursion. Let's transform it into something more manageable, ie.~of finite arity.

@@

We can also move \(T(n)\) to the other side, ie.:

@@

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \((2 n - 1) T(n)\) describes well-weighted binary plane trees of size \(n\) (ie. with \(n\) leaves) with some node pointed (internal or leaf),
    \item \((n + 1) T(n + 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some leaf pointed,
    \item \((n - 2) T(n - 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some internal node pointed.
\end{itemize}

@@

First thing that we have to select is defining the size carefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (ie.\ children of the leaves),
    \item counting edges.
\end{itemize}

@@

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, eg.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common roots and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates a tree of size being the sum of sizes of the components increased by 1. If you had defined the size as the number of leaves or null-nodes, the size would be just a simple sum of the sizes of the smaller trees.
\end{itemize}

@@

If we describe some structure we may be tempted to use multiple kinds of nodes, eg.~unary and binary nodes. Having multiple kinds of nodes may cause problems, both in creating holonomic recurrence and finding the interpretation, as we have to keep track of how they can transform.

@@

Most of the time we are operating on trees with some nodes of weight \(0\) and some of weight \(1\). We may as well try to make them variable or assign some other values (eg.\ small prime numbers).

@@

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such a simple case as binary trees is almost trivial, especially if one has seen the solution once.

@@

\begin{definition}
    Set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a variable, \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (application of function \(M\) to an argument \(N\))
        \item if \(x\) is a variable and \(M \in \Lambda\), then \((\lambda x . M) \in \Lambda\) (abstraction, ie.\ creating a function with an argument \(x\) and binding each occurrence of \(x\) in \(M\) to this argument)
    \end{itemize}
\end{definition}

The definition above is quite popular, but introduces several problems. One of them is a problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable bound (or is it even a free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

@@

In order to make the image cleaner, we will drop the arguments of generating functions, ie. \(\gf{L} := \gf{L}(z, o, l, a, s)\)

@@

You may notice that some potentially trivial terms are more complex than they seem. If we didn't count different types of nodes, we would achieve following holonomic recurrence:

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate between nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, ie.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, ie.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, ie.~is right child of its parent;
    \item node has its parent on the right, ie.~is the left child of its parent.
\end{itemize}

@@

\begin{definition}
    In the context of \(T_i\), \(C_k\) is defined as \(L(n - i)\), ie.\ number of tree transformations from \(T_i\) are working on.
\end{definition}

@@

We are going to generate two kinds of trees. Positive and negative ones. When computing the total number of trees, we count positive trees as \(1\), negative as \(-1\). Our goal is to generate \(E_{i, j}\) trees for each section \(T_{i, j}\).

@@

c.append_r_class_subtree_visitor(1,       # we define positive trees
                                          # ie. that count as 1
    lambda kind, left, right:
        None if kind[0] is not None else  # we ignore all actual nodes
        [kind + ("*", )] + left + right   # for each virtual node we point it
)
\end{lstlisting}

@@

Basically, for each virtual node not being root (ie.\ having left or right parent), we point it and add \(l\) or \(s\) respectively above the parent.

@@

The process can then be repeated for each section. We have chosen lexicographical order (ie. \(0,0\); \(1,0\); \(1,1\); \(2,0\); \(\ldots\)).

@@

You may also note that we didn't generate all trees suggested by the recurrence. This is due to one cheat we used.

@@

You may remember our assumptions in section \ref{sub:the_framework} about the discrimination cases. We allowed ourselves to apply visitors only to one of the left and right children, which isn't handled by our holonomic recurrence as the number of such vertices does not depend only on \(n\).

@@

The framework also isn't perfect. It highly depends on the operator's intuition. We believe that the solving process can be automated. The simplest method that comes to mind is simply creating heuristic rules that then will be used by the framework to guess at least partial solutions.

>>>>

Every recursive structure can be described using a generating function, which can be then transformed into holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations, like adding some vertices to the smaller trees. Although, there is no known method of interpreting it in the language of the combinatorial structures.

@@

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an approach that helped us to find an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation.

@@

Section\ \ref{sec:brutal_iterative_approach} presents our final approach to the original problem, i.e.\ computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is able to provide us with the full solution, but we were not able to achieve that due to the limited time that was available to us.

@@

In order to make reading this thesis easier, let's fix some notation and make some assumptions. Any deviation from this will be clearly stated. If you have trouble guessing what some letter or word means, this should be your primary place to look for the answer.

@@

\(\no\) is implicitly defined as an identifier of a non-existing node (e.g.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

Dashed edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__dashed_edge}
\end{minipage}
depict edges that may or may not exist (e.g.\ when we describe transformations and we do not care if the node is root or if it has children).

@@

Virtual node is an artificially defined node that does not affect the structure definition (e.g.\ we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

@@

e.g.\ stands for Latin exempli gratia, which means \textit{for example}.

i.e.\ stands for Latin id est, which means \textit{that is}.
\todo{It should be obvious, but people mix them up}

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define a full binary tree recursively as a node having no children (i.e.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

Throughout the thesis we use ordinary generating functions to describe and count structures.

@@

The highlighted terms describe following tree:

@@

\begin{definition}[\cite{holotoolkit}]
    \label{def:holo_1}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) there exists a holonomic recurrence for it, i.e.:

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, i.e.\ none of them satisfies a linear recurrence equation with polynomial coefficients\cite{nonholo}.
\end{example}

\begin{definition}[\cite{complexity}]
    \label{def:holo_2}
    Equivalently to definition\ \ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) there exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identical to zero.
\end{definition}

@@

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, e.g.\ we may want to find the number of trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent.

@@

In the thesis we try to gather methods and results that may be helpful for anyone trying to find such interpretation, as we believe that such interpretation may give better insight into the nature of the structure and make it easier to study its behaviour.

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as:
    \[\enc{T} = \enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\]
    i.e.\ catenation of the identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

@@

Throughout thesis we define some operations using this notation. As pre-order notation is bijective, you can obtain the structure the notation describes by reversing the process of encoding.

@@

We define the size of a tree as a number of internal nodes (i.e.\ nodes that are not leaves). Our generating function is therefore:

@@

It is easy to see that a tree of size \(n\) (i.e.\ with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

Let's define two local operations: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of tree with predefined size (e.g.\ node and its 2 children, but can not traverse the whole tree).

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.\ trees of size \(n\) with a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\)

@@

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary trees. We do not change the structure, just the notation.

@@

The bijection between unary-binary and left-leaning binary trees is very simple. We just represent unary nodes as nodes with just the left child of the left-leaning binary tree. The other cases translate immediately to corresponding ones. Please note that this transformation keeps the size of the tree (i.e.~number of edges).

@@

All we need to show are the transformations from the smaller cases, i.e.\ \(U(n-1)\) and \(U(n-2)\).

@@

This definition is quite precise, but is difficult to implement as a structural recursion. Let's transform it into something more manageable, i.e.~of finite arity.

@@

We can also move \(T(n)\) to the other side, i.e.:

@@

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \((2 n - 1) T(n)\) describes well-weighted binary plane trees of size \(n\) (i.e.\ with \(n\) leaves) with some node pointed (internal or leaf),
    \item \((n + 1) T(n + 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some leaf pointed,
    \item \((n - 2) T(n - 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some internal node pointed.
\end{itemize}

@@

First thing that we have to select is defining the size carefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (i.e.\ children of the leaves),
    \item counting edges.
\end{itemize}

@@

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, e.g.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common roots and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates a tree of size being the sum of sizes of the components increased by 1. If you had defined the size as the number of leaves or null-nodes, the size would be just a simple sum of the sizes of the smaller trees.
\end{itemize}

@@

If we describe some structure we may be tempted to use multiple kinds of nodes, e.g.~unary and binary nodes. Having multiple kinds of nodes may cause problems, both in creating holonomic recurrence and finding the interpretation, as we have to keep track of how they can transform.

@@

Most of the time we are operating on trees with some nodes of weight \(0\) and some of weight \(1\). We may as well try to make them variable or assign some other values (e.g.\ small prime numbers).

@@

Using our inductive hypothesis, i.e.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such a simple case as binary trees is almost trivial, especially if one has seen the solution once.

@@

\begin{definition}
    Set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a variable, \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (application of function \(M\) to an argument \(N\))
        \item if \(x\) is a variable and \(M \in \Lambda\), then \((\lambda x . M) \in \Lambda\) (abstraction, i.e.\ creating a function with an argument \(x\) and binding each occurrence of \(x\) in \(M\) to this argument)
    \end{itemize}
\end{definition}

The definition above is quite popular, but introduces several problems. One of them is a problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable bound (or is it even a free variable, i.e.\ not bound). To address that, we can use de Bruijn indices.

@@

In order to make the image cleaner, we will drop the arguments of generating functions, i.e.\ \(\gf{L} := \gf{L}(z, o, l, a, s)\)

@@

You may notice that some potentially trivial terms are more complex than they seem. If we did not count different types of nodes, we would achieve following holonomic recurrence:

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate between nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, i.e.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, i.e.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, i.e.~is right child of its parent;
    \item node has its parent on the right, i.e.~is the left child of its parent.
\end{itemize}

@@

\begin{definition}
    In the context of \(T_i\), \(C_k\) is defined as \(L(n - i)\), i.e.\ number of tree transformations from \(T_i\) are working on.
\end{definition}

@@

We are going to generate two kinds of trees. Positive and negative ones. When computing the total number of trees, we count positive trees as \(1\) and negative as \(-1\). Our goal is to generate \(E_{i, j}\) trees for each section \(T_{i, j}\).

@@

c.append_r_class_subtree_visitor(1,       # we define positive trees
                                          # i.e. that count as 1
    lambda kind, left, right:
        None if kind[0] is not None else  # we ignore all actual nodes
        [kind + ("*", )] + left + right   # for each virtual node we point it
)
\end{lstlisting}

@@

Basically, for each virtual node not being root (i.e.\ having left or right parent), we point it and add \(l\) or \(s\) respectively above the parent.

@@

The process can then be repeated for each section. We have chosen lexicographical order (i.e.\ \(0,0\); \(1,0\); \(1,1\); \(2,0\); \(\ldots\)).

@@

You may also note that we did not generate all trees suggested by the recurrence. This is due to one cheat we used.

@@

You may remember our assumptions in section \ref{sub:the_framework} about the discrimination cases. We allowed ourselves to apply visitors only to one of the left and right children, which is not handled by our holonomic recurrence as the number of such vertices does not depend only on \(n\).

@@

The framework also is not perfect. It highly depends on the operator's intuition. We believe that the solving process can be automated. The simplest method that comes to mind is simply creating heuristic rules that then will be used by the framework to guess at least partial solutions.
